A molecular dynamics study of the solvation of carbon dioxide in the ionic liquids \ce{[emim][B(CN)_4]} and \ce{[emim][NTf_2]}

### 2018-09-07 ionic_liquids.tex
Refinement of Theoretical Background section

In this section, we present the formalism that allows us to compute solvation free energies from molecular dynamics simulations and, after that, obtain Henry constants and activity coefficients.
For dealing with systems at infinite dilution, we rely on the work of Shirts \textit{et al}. \cite{Shirts_2003}.
Then, we reformulate the equations in order to study concentrated solutions, which is a more likely scenario in the case of \ce{CO_2} capture processes.

@@

The solvation free energy, $\Delta G_{\text{solv}}$, is defined as the reversible work needed to transfer a solute molecule from a gas phase to the solution.
For dense systems, an efficient computation of solvation properties via molecular simulation is possible by means of the so-called alchemical methods, which employ non-physical pathways to reversibly connect the endpoint states whose difference in free energy is to be estimated.
In the case of $\Delta G_{\text{solv}}$, the alchemical route involves a gradual insertion/deletion of the solute through the use of a coupling parameter $\lambda$, which scales the solute-solvent interactions by taking values in the interval [0,1]. By convention, $\lambda = 1$ corresponds to a fully interacting system, while for $\lambda$ = 0 the solute-solvent interactions are turned off completely.

In what follows, the subscripts ``s'' and ``IL'' refer to the solute and the solvent (ionic liquid), respectively.
For a liquid phase containing $N_\text{\tiny IL}$ solvent molecules and $N_{\text{s}}$ solute molecules at fixed pressure (P) and temperature (T), the difference in free energy between the states associated to $\lambda$ = 1 and $\lambda$ = 0 is formally defined as
\begin{equation}
\Delta G_\text{sim} \equiv k_b T \ln \frac{\Xi (N_s,N_\text{\tiny IL},P,T,\lambda = 1)}{\Xi (N_s,N_\text{\tiny IL},P,T,\lambda = 0)}, 
\end{equation}
where $\Xi$ denotes the isobaric-isothermal partition function and $k_b$ is the Boltzmann constant.
Shirts \textit{et al}. \cite{Shirts_2003} rigorously derived the working expression that relates $\Delta G_\text{sim}$ to what is experimentally measured as the solvation free energy, namely
\begin{equation}
\label{eq:solvation free energy definition}
\Delta G_{\text{solv}} \equiv k_b T \ln \left( \frac{P_s}{c_s R T} \right),
\end{equation}
where $R = N_A k_b$ is the gas constant,
$N_A$ is the Avogadro number,
$P_s$ is the partial pressure of the solute in the gas phase, and
$c_s$ is the molar concentration of the solute in the liquid phase.
Note that these quantities correspond to a phase equilibrium condition.
For obtaining the mentioned relation, the authors derived an expression for the chemical potential of the solute in the liquid phase ($\mu^l_s$) as a function of $\Delta G_\text{sim}$, which can be written as
\begin{equation}
\label{eq:solute chemical potential}
\mu^l_s = \mu^{ig}_s + \Delta G_\text{sim} - k_bT \ln \left( \frac{V^{\ast}}{V_1}\right) - k_bT \ln \left( \frac{P_s}{c_s R T} \right),
\end{equation}
where $\mu^{ig}_s$ is the chemical potential of the solute in an ideal gas at temperature $T$ and solute partial pressure $P_s$, while $V^{\ast}$ and $V_1$ are the mean volumes of liquid solutions specified by ($N_s-1$,$N_\text{\tiny IL}$,P,T) and ($N_s$,$N_\text{\tiny IL}$,P,T), respectively.
Note that $c_s = \frac{N_s}{N_A V_1}$.
For small solutes, such as those considered in this work, the term containing the ratio $V^{\ast}/V_1$ is typically in the order of magnitude of the free-energy uncertainties \cite{Shirts_2003}.
Hence, it can be neglected and will be omitted hereafter.

Finally, by treating the gas phase as an ideal gas and considering the system to be at equilibrium ($\mu^l_s = \mu^{ig}_s$), Shirts \textit{et al}. \cite{Shirts_2003} found out that
\begin{equation}
\begin{split}
\label{eq:free_solv}
 \Delta G_{\text{solv}} \approx \Delta G_\text{sim}.
\end{split}
\end{equation}

@@

If both $P_s$ and $c_s$ are small, we can resort to Eqs.~\eqref{eq:solvation free energy definition} and \eqref{eq:free_solv} to predict Henry's law constant $\text{K}_s$, which is defined as \cite{Prausnitz}
\begin{equation}
\text{K}_s =\frac{P_s}{x_s}.
\end{equation}

Therefore, by simply expressing Eq.~\eqref{eq:solvation free energy definition} in terms of the molar fraction of the solute in the liquid phase ($x_s$) and assuming a dilute solution, whose molar density does not perceivably differ from that of pure solvent (that is, $\tilde{\rho} \approx \tilde{\rho}_\text{\tiny IL}$), we obtain that
\begin{equation}
\label{eq:henry_eq}
K_s = \tilde{\rho}_\text{\tiny IL} R \, T \exp\left(\frac{\Delta G_\text{sim}}{k_b T}\right).
\end{equation}

@@

For the ILs of interest to the present investigation, high concentrations of \ce{CO_2} in the liquid phase are obtained at pressure values for which, in principle, the ideal gas assumption may not be valid.
In this case, we can compute the chemical potential of the solute in the real gas phase as
\begin{equation}
\label{eq:mu_gas_real}
\mu^{g}_s = \mu^{ig}_s + k_b T \ln \left(\frac{f^{g}_s}{P_s}\right),
\end{equation}
where $f^{g}_s$ is the fugacity of the solute in the gas phase.
Then, by considering a system at phase equilibrium ($\mu^{l}_s$ = $\mu^{g}_s$), it follows that
\begin{equation}
\label{eq:fgas_d}
f^{g}_s = \exp \left( \frac{\Delta G_\text{sim}}{k_b T} \right) +  c_s R T. 
\end{equation}

Three additional assumptions are made here. First, we assume that the gas phase is either pure solute (which is consistent with the ionic-liquid systems with which we compare our simulations) or can be approximated by an ideal solution. Second, we assume that the system pressure is moderate, so that a truncated virial equation is enough to represent the gas phase. In this case, Eq.~\eqref{eq:fgas_d} becomes
\begin{equation}
\label{eq:equation for pressure}
y_s P \exp\left( \frac{B_s P}{R T} \right) = \exp \left( \frac{\Delta G_\text{sim}}{k_b T} \right) +  c_s R T,
\end{equation}
where $y_s$ is the molar fraction of solute in the gas phase ($y_s \rightarrow 1$ for the systems we analyze here) and $B_s$ is the second virial coefficient of the solute.
In the case of \ce{CO_2}, we employ the correlation for $B_s$ developed by Holste \textit{et al}. \cite{Holste_1987}, which is
\begin{equation}
B_s = B_0 + \frac{B_1}{T} + \frac{B_2}{T^2},
\end{equation}
where $B_0 = 23.02991~\mathrm{\frac{cm^3}{mol}}$, $B_1 = -2455.297~\mathrm{\frac{K \cdot cm^3}{mol}}$, and $B_2 = -1.22675 \times 10^7~\mathrm{\frac{K^2 \cdot cm^3}{mol}}$.

Our third assumption, supported by preliminary simulations, is that $\Delta G_\text{sim}$ does not vary appreciably with pressure, at least over the range of experimental data considered here.
In this way, no iterative simulations are required to compute the equilibrium pressure by solving Eq.~\eqref{eq:equation for pressure} for $P$.

@@

As will be shown below, the prediction of activity coefficients requires additional simulations of the solute as a pure liquid phase, whose properties we denote here by a superscript ``$0$''.
We start by using standard thermodynamic relations \cite{Tester} to write the solute activity coefficient $\gamma_s$ in terms of chemical potentials as
\begin{equation}
\label{eq:gamma}
\ln \gamma_s = \frac{\mu_s^l - \mu_s^{l,0}}{k_B T} - \ln x_s,
\end{equation}

Next, we replace $\mu_s^l$ and $\mu_s^{l,0}$ by their corresponding instances of Eq.~\eqref{eq:solute chemical potential}. In this step, we note that $c_s = \tilde{\rho} x_s$ and $c_s^0 = \tilde{\rho}_s^0$, where $\tilde{\rho}$ and $\tilde{\rho}_s^0$ are the molar densities of the solution and of the pure liquid solute, respectively. Thus, the expression we obtain for computing an activity coefficient from two independent simulations is
\begin{equation}
\ln \gamma_s = \frac{\Delta G_\text{sim} - \Delta G_\text{sim}^{0}}{k_b T} + \ln \frac{\tilde{\rho}}{\tilde{\rho}_s^0}.
\end{equation}

@@

\section{Force fields}
\label{sec:force_field}
\subsection{Ionic liquids}
\label{sec:force_field_il}

>>>>

In this section, we present the formalism that allows us to compute solvation free energies from molecular dynamics simulations and, after that, to obtain Henry constants and activity coefficients.
For dealing with systems at infinite dilution at low pressures, we rely on the work of Shirts \textit{et al}. \cite{Shirts_2003}.
Then, we reformulate the equations in order to study concentrated solutions and systems at higher pressures, which is a more likely scenario in the case of \ce{CO_2} capture processes.

@@

The solvation free energy, $\Delta G_{\text{solv}}$, is defined as the reversible work needed to transfer a solute molecule from a gas phase to a liquid solution.
For dense systems, an efficient computation of solvation properties via molecular simulation is possible by means of the so-called alchemical methods, which employ non-physical pathways to reversibly connect the endpoint states whose difference in free energy is to be estimated.
In the case of $\Delta G_{\text{solv}}$, the alchemical route involves a gradual insertion/deletion of the solute through the definition of a coupling parameter $\lambda$, which scales the solute-solvent interactions by taking values in the interval [0,1]. By convention, $\lambda = 1$ corresponds to a fully interacting system, while for $\lambda$ = 0 the solute-solvent interactions are turned off completely.

In what follows, the subscripts ``s'' and ``IL'' refer to the solute and the solvent (ionic liquid), respectively.
For a liquid phase containing $N_\text{\tiny IL}$ solvent molecules and $N_{\text{s}}$ solute molecules at fixed pressure (P) and temperature (T), the difference in free energy between the states associated to $\lambda$ = 1 and $\lambda$ = 0 is formally defined as
\begin{equation}
\Delta G_\text{sim} \equiv k_b T \ln \frac{\Xi (N_s,N_\text{\tiny IL},P,T,\lambda = 1)}{\Xi (N_s,N_\text{\tiny IL},P,T,\lambda = 0)}, 
\end{equation}
where $\Xi$ denotes the isobaric-isothermal partition function and $k_b$ is the Boltzmann constant.
Shirts \textit{et al}. \cite{Shirts_2003} rigorously derived the working expression that relates $\Delta G_\text{sim}$ to what is experimentally measured as the solvation free energy, namely
\begin{equation}
\label{eq:solvation free energy definition}
\Delta G_{\text{solv}} \equiv k_b T \ln \left( \frac{P_s}{c_s R T} \right),
\end{equation}
where $R = N_A k_b$ is the gas constant,
$N_A$ is the Avogadro number,
$P_s$ is the partial pressure of the solute in the gas phase, and
$c_s$ is the molar concentration of the solute in the liquid phase.
Note that these quantities correspond to a phase equilibrium condition.
For obtaining the mentioned relation, the authors derived an expression for the chemical potential of the solute in the liquid phase ($\mu^l_s$) as a function of $\Delta G_\text{sim}$, which can be written as
\begin{equation}
\label{eq:solute chemical potential}
\mu^l_s = \mu^{ig}_s + \Delta G_\text{sim} - k_bT \ln \left( \frac{V^{\ast}}{V_1}\right) - k_bT \ln \left( \frac{P_s}{c_s R T} \right),
\end{equation}
where $\mu^{ig}_s$ is the chemical potential of the solute in an ideal gas at temperature $T$ and solute partial pressure $P_s$, while $V^{\ast}$ and $V_1$ are the mean volumes of liquid solutions specified by ($N_s-1$,$N_\text{\tiny IL}$,P,T) and ($N_s$,$N_\text{\tiny IL}$,P,T), respectively.
Note that $c_s = \frac{N_s}{N_A V_1}$.

Finally, by treating the gas phase as an ideal gas and considering the system to be at equilibrium ($\mu^l_s = \mu^{ig}_s$), Shirts \textit{et al}. \cite{Shirts_2003} found out that
\begin{equation}
\begin{split}
\label{eq:free_solv}
 \Delta G_{\text{solv}} = \Delta G_\text{sim} - k_bT \ln \left( \frac{V^{\ast}}{V_1}\right).
\end{split}
\end{equation}

For small solutes, such as those considered in this work, the term containing the ratio $V^{\ast}/V_1$ is typically in the order of magnitude of the free-energy uncertainties \cite{Shirts_2003}.
Hence, it can be neglected and will be omitted hereafter.

@@

If both $P_s$ and $c_s$ are small, we can resort to Eqs.~\eqref{eq:solvation free energy definition} and \eqref{eq:free_solv} to predict Henry's law constant, which is defined as $\text{K}_s =\frac{P_s}{x_s}$ \cite{Prausnitz}.
%\begin{equation}
%\text{K}_s =\frac{P_s}{x_s}.
%\end{equation}
%
Therefore, by simply expressing Eq.~\eqref{eq:solvation free energy definition} in terms of the molar fraction of the solute in the liquid phase ($x_s$) and assuming a dilute solution, whose molar density does not perceivably differ from that of pure solvent (that is, $\tilde{\rho} \approx \tilde{\rho}_\text{\tiny IL}$), we obtain that
\begin{equation}
\label{eq:henry_eq}
K_s = \tilde{\rho}_\text{\tiny IL} R \, T \exp\left(\frac{\Delta G_\text{sim}}{k_b T}\right).
\end{equation}

@@

For the ILs of interest to the present investigation, high concentrations of \ce{CO_2} in the liquid phase are obtained at pressure values for which, in principle, the ideal gas assumption might no longer be valid.
In this case, we can compute the chemical potential of the solute in the real gas phase as
\begin{equation}
\label{eq:mu_gas_real}
\mu^{g}_s = \mu^{ig}_s + k_b T \ln \left(\frac{f^{g}_s}{P_s}\right),
\end{equation}
where $f^{g}_s$ is the fugacity of the solute in the gas phase.
Then, by considering a system at phase equilibrium ($\mu^{l}_s$ = $\mu^{g}_s$), it follows that
\begin{equation}
\label{eq:fgas_d}
f^{g}_s = \exp \left( \frac{\Delta G_\text{sim}}{k_b T} \right) +  c_s R T. 
\end{equation}

Three additional assumptions are made here. First, we assume that the gas phase is either pure solute (which is consistent with the ionic-liquid systems with which we compare our simulations) or can be approximated by an ideal solution. Second, we assume that the system pressure is moderate, so that a truncated virial equation is enough to represent the gas phase. In this case, Eq.~\eqref{eq:fgas_d} becomes
\begin{equation}
\label{eq:equation for pressure}
y_s \exp\left( \frac{B_s P}{R T} \right) P = \exp \left( \frac{\Delta G_\text{sim}}{k_b T} \right) +  c_s R T,
\end{equation}
where $y_s$ is the molar fraction of solute in the gas phase ($y_s \rightarrow 1$ for the systems we analyze here) and $B_s$ is the second virial coefficient of the solute.
In the case of \ce{CO_2}, we employ the correlation for $B_s$ developed by Holste \textit{et al}. \cite{Holste_1987}, which is
\begin{equation}
B_s = B_0 + \frac{B_1}{T} + \frac{B_2}{T^2},
\end{equation}
where $B_0 = 23.02991~\mathrm{\frac{cm^3}{mol}}$, $B_1 = -2455.297~\mathrm{\frac{K \cdot cm^3}{mol}}$, and $B_2 = -1.22675 \times 10^7~\mathrm{\frac{K^2 \cdot cm^3}{mol}}$.

Our third assumption, supported by preliminary simulations, is that $\Delta G_\text{sim}$ does not vary appreciably with pressure, at least over the range of experimental data considered here.
In this way, no iterative simulations are required to compute the equilibrium pressure when solving Eq.~\eqref{eq:equation for pressure} for $P$.

@@

As will be shown below, the prediction of activity coefficients requires additional simulations of the solute as a pure liquid phase, whose properties we denote here by a superscript ``$0$''.
We start by using a standard thermodynamic relation \cite{Tester} to write the solute activity coefficient $\gamma_s$ in terms of chemical potentials as
\begin{equation}
\label{eq:gamma}
\ln \gamma_s = \frac{\mu_s^l - \mu_s^{l,0}}{k_B T} - \ln x_s,
\end{equation}

Next, we replace $\mu_s^l$ and $\mu_s^{l,0}$ by their corresponding instances of Eq.~\eqref{eq:solute chemical potential}. In this step, we note that $c_s = \tilde{\rho} x_s$ and $c_s^0 = \tilde{\rho}_s^0$, where $\tilde{\rho}$ and $\tilde{\rho}_s^0$ are the molar densities of the solution and of pure liquid solute, respectively. Thus, the expression we obtain for computing an activity coefficient from two independent simulations is
\begin{equation}
\ln \gamma_s = \frac{\Delta G_\text{sim} - \Delta G_\text{sim}^{0}}{k_b T} + \ln \frac{\tilde{\rho}}{\tilde{\rho}_s^0}.
\end{equation}

@@

\section{Force fields}
\label{sec:force_field}

\subsection{Ionic liquids}
\label{sec:force_field_il}
